Deployable Probabilistic Programming

### 2019-04-23 paper/paper.tex
empirical

\linespread{1.1}

@@

Python~\cite{Python95}, Scala~\cite{Scala06}, and
Julia~\cite{Julia14} have been popular choices for development
of probabilistic programming languages and frameworks integrated
with general-purpose programming
languages~\cite{S16,MMW18,BCJ+19,P09,Rainier,GXG18,ISF+18}.
Each of these languages has advantages for implementing a
probabilistic programming facility. However, for a reference
implementation of a facility best suited for probabilistic
inference in a porduction software system, we were looking
for a language that \begin{itemize}
  \item compiles into fast and compact code;
  \item does not incur dependency on a heavy runtime; and
  \item is a popular choice for building server-side software.
\end{itemize}
In addition, since we were going to automatically differentiate
the language's source code, we were looking for a relatively
small language, preferably one for which parsing and program
analysis tools are readily available.

@@

\begin{table*}
	\begin{tabular}{r | c |  c | c | c | c | c | }
		{\it model}   & \multicolumn{6}{c|}{\it time, seconds} \\
					  & \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
					  & Infergo & Turing & Stan              & Infergo & Turing & Stan    \\\hline
		Eight schools & 0.5     &        & 50                & 0.6     &        & 0.12    \\
		GMM           & 0.5     &        & 55                & 3.0     &        & 4.9     \\
		LDA           & 0.5     &        & 55                &         &        & 3.7     \\ 
	\end{tabular}
\caption{Compilation and execution times for 1000 iterations 
	of HMC with 10 leapfrog steps.}
\label{tab:memory-runtime}
\end{table*}

Table~\ref{tab:memory-runtime} provides memory consumption and
running time measurements on basic models to illustrate
Infergo's performance.  The measurements were obtained on a
2.3GHz Intel Core 5 CPU with 8GB of memory for 1000 iterations
of Hamiltonian Monte Carlo with 10 leapfrog steps. Note that
log-likelihood computation for standard distributions is not
optimized yet. Quite the opposite: since models in Infergo are
fully composable, primitive distributions are themselves
implemented as Infergo models and automatically differentiated.

@@

Building a production software system which employs
probabilistic inference as an integral part of prediction and
decision making algorithms is challenging. To address the
challenges, we proposed guidelines for implementation of a
deployable probabilistic programming facility. Further on, we
introduced Infergo, a probabilistic programming facility in Go,
as a reference implementation in accordance with the guidelines.
Infergo allows to program inference models in virtually
unrestricted Go, and to perform inference on the model using
either supplied or third-party algorithms. We demonstrated on
case studies how Infergo overcomes the challenges by following
the guidelines.

@@

>>>>

\linespread{1.07}

@@

Python~\cite{Python95}, Scala~\cite{Scala06}, and
Julia~\cite{Julia14} have been popular choices for development
of probabilistic programming languages and frameworks integrated
with general-purpose programming
languages~\cite{S16,MMW18,BCJ+19,P09,Rainier,GXG18,ISF+18}.
Each of these languages has advantages for implementing a
probabilistic programming facility. However, for a reference
implementation of a facility best suited for probabilistic
inference in a production software system, we were looking
for a language that \begin{itemize}
  \item compiles into fast and compact code;
  \item does not incur dependency on a heavy runtime; and
  \item is a popular choice for building server-side software.
\end{itemize}
In addition, since we were going to automatically differentiate
the language's source code, we were looking for a relatively
small language, preferably one for which parsing and program
analysis tools are readily available.

@@

\begin{table*}
	\begin{tabular}{r | c |  c | c | c | c | c | }
		{\it model}         & \multicolumn{6}{c|}{\it time, seconds} \\
					        & \multicolumn{3}{c|}{\it compilation} & \multicolumn{3}{c|}{\it execution} \\
					        & Infergo & Turing & Stan              & Infergo  & Turing & Stan    \\\hline
              Eight schools & 0.50    & -      & 50                & 0.60     & 2.8    & 0.12    \\
     Gaussian mixture model & 0.50    & -      & 54                & 32       & 14     & 4.9     \\
Latent Dirichlet allocation & 0.50    & -      & 54                & 8.9      & 12     & 3.7     \\ 
	\end{tabular}
\caption{Compilation and execution times for 1000 iterations 
	of HMC with 10 leapfrog steps.}
\label{tab:memory-runtime}
\end{table*}
It is customary to include performance evaluation on benchmark
problems in publications on new probabilistic programming
frameworks~\cite{PW14,WVM14,TMY+16,GXG18}. We provide here a
comparison of running times of Infergo and two other probabilistic
programming frameworks: Turing~\cite{GXG18} and
Stan~\cite{Stan17}. Turing is a Julia~\cite{Julia14} library
and DSL offering exploratory statistical modeling  and
inference composition. Stan is a compiled language and a library
optimized for high performance.

We use three models of different size and structure for the
comparison: two of the models, Eight schools and Gaussian
mixture model, were introduced as a part of the case
studies~(Section~\ref{sec:case-studies}); the code for the third
model, Latent Dirichlet allocation~\cite{BNG03}, is included in
Appendix~\ref{app:lda}. The code and data for all models were
originally included in Stan examples and translated for use
with Turing and Infergo. \textit{Eight schools} is the smallest model.
One may expect that most of the time execution time is spent in
executing the boilerplate code rather than the model. \textit{Gaussian
mixture model} is a more elaborated model which benefits greatly
from vectorization. The greater the number of observations (1000
data points were used in the evaluation), the more significant
is the benefit of vectorized computations. \textit{Latent
Dirichlet allocation} is, again, a relatively elaborated model,
that is, however, not as easy to vectorize.

We report both compilation time (for Stan and Infergo) and
execution time. Compilation time is an important characteristic
of a probabilistic programming framework --- statistical
modeling often involves many cycles of model modifications and
numerical experiments on subsampled data. The ability to modify
and instantly re-run a model is crucial for exploratory data
analysis and model design. Turing is based on Julia, which is
a dynamic language with JIT compiler, and there is no separate
compilation time. For Stan, the compilation time includes
compiling the Stan model to C++, and then compiling the C++
source into executable code for inference. For Infergo, the
compilation time consists of automatic differentiation of the
model, and then building a command-line executable that calls
an inference algorithm on the model.

Table~\ref{tab:memory-runtime} shows running time measurements
on the models.  The measurements were obtained on a 1.25GHz
Intel(R) Core(TM) i5 CPU with 8GB of memory for 1000 iterations
of Hamiltonian Monte Carlo with 10 leapfrog steps and averaged
over 10 runs.  In execution, Infergo is slower than Stan, which
comes as little surprise --- Stan is a mature highly optimized
inference and optimization library. However, Infergo and Turing
show similar execution times, despite Turing relying on Julia's
numerical libraries. The slowest relative execution time is on
the Gaussian mixture model, where Stan is 6--7 times faster,
apparently because Infergo model is not vectorized; the model
could be made more efficient at the cost of verbosity and poorer
readability, but we opted to use idiomatic Infergo models for
the comparison. On the Latent Dirichlet allocation model, run on
Stan example data with 25 documents and 262 word instances,
Infergo is only 2.5 times slower than Stan, and faster than
Turing, an illustration of the benefits of transformation-based
differentiation and efficiency of Go.

Infergo programs compile very fast, in particular in comparison
to Stan models. Both automatic differentiation and code
generation combined take a fraction of a second. Compilation time
of Stan models is dominated by compilation of the generated C++
code which takes close to a minute on our hardware. The ability
to modify a model and then re-run inference almost instantly,
and with reasonable performance, helps in exploratory model
development with Infergo, which was conceived with server-side,
unattended execution in mind, but fits the exploration and
development stage just as well.

@@

Building a production software system which employs
probabilistic inference as an integral part of prediction and
decision making algorithms is challenging. To address the
challenges, we proposed guidelines for implementation of a
deployable probabilistic programming facility. Based on the
guidelines, we introduced Infergo, a probabilistic programming
facility in Go, as a reference implementation in accordance with
the guidelines.  Infergo allows to program inference models in
virtually unrestricted Go, and to perform inference on the model
using either supplied or third-party algorithms. We demonstrated
on case studies how Infergo overcomes the challenges by
following the guidelines.

@@

\clearpage

\appendix

\begin{figure*}
\section{Latent Dirichlet Allocation Model}
	\label{app:lda}
\vspace{\baselineskip}

\begin{minipage}{0.49\textwidth}
\centering
\textbf{Infergo}

\begin{lstlisting}[framexleftmargin=10pt,numbers=left]
type Model struct {
  K     int       // num topics
  V     int       // num words
  M     int       // num docs
  N     int       // total word instances
  Word  []int     // word n
  Doc   []int     // doc for word n
  Alpha []float64 // topic prior
  Beta  []float64 // word prior
}

func (m *Model) Observe(x []float64) float64 {
  ll := 0.0

  // Destructure parameters
  theta := make([][]float64, m.M)
  m.simplices(&x, m.K, theta)
  phi := make([][]float64, m.K)
  m.simplices(&x, m.V, phi)

  // Impose priors
  ll += Dirichlet{m.K}.Logps(m.Alpha, theta...)
  ll += Dirichlet{m.V}.Logps(m.Beta, phi...)

  // Conditioning on observations
  gamma := make([]float64, m.K)
  for in := 0; in != m.N; in++ {
    for ik := 0; ik != m.K; ik++ {
      gamma[ik] = math.Log(theta[m.Doc[in]-1][ik]) +
        math.Log(phi[ik][m.Word[in]-1])
    }
    ll += D.LogSumExp(gamma)
  }
  return ll
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.49\textwidth}
	\centering

	\textbf{Stan}

\begin{lstlisting}[language=Stan]
data {
  int<lower=2> K;               // num topics
  int<lower=2> V;               // num words
  int<lower=1> M;               // num docs
  int<lower=1> N;           // total word instances
  int<lower=1,upper=V> w[N];    // word n
  int<lower=1,upper=M> doc[N];  // doc for word n
  vector<lower=0>[K] alpha;     // topic prior
  vector<lower=0>[V] beta;      // word prior
}

parameters {
  simplex[K] theta[M]; // topic dist for doc m
  simplex[V] phi[K];   // word dist for topic k
}

model {
  for (m in 1:M)  
    theta[m] ~ dirichlet(alpha);  // prior
  for (k in 1:K)  
    phi[k] ~ dirichlet(beta);     // prior

  for (n in 1:N) {
    real gamma[K];
    for (k in 1:K) 
      gamma[k] <- log(theta[doc[n],k]) 
	  	+ log(phi[k,w[n]]);
    increment_log_prob(log_sum_exp(gamma));
  }
}
\end{lstlisting}
	\vspace{2\baselineskip}
\end{minipage}
\end{figure*}
